A Simple Type Theory for Category Theory

### 2022-02-05 paper/lipics-v2021-sample-article.tex
first subsection of syntax :$

\usepackage[margin=1in]{geometry}
\usepackage{tikz-cd}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{stmaryrd}
\usepackage{mathpartir}
\usepackage{bbm}
\usepackage{cleveref}

@@

\begin{document}
%% \newtheorem{theorem}{Theorem}
%% \newtheorem{lemma}{Lemma}
%% \newtheorem{construction}{Construction}
%% \newtheorem{definition}{Definition}
\newcommand{\pto}{\nrightarrow}
\newcommand{\pfrom}{\nleftarrow}
\newcommand{\vcat}{\mathcal}
\newcommand{\cat}{\mathbbm}
\newcommand{\isaSet}{\,\,\textrm{Set}}
\newcommand{\Set}{\textrm{Set}}
\newcommand{\vtkmnd}{\mathbb{K}\text{Mod} (\vcat{V},T)}
\newcommand{\rmod}{\text{RMod}}
\newcommand{\lmod}{\text{LMod}}
\newcommand{\varr}[2]{\text{Fun}{#1}{#2}}
\newcommand{\harr}[2]{\text{Rel}{#1}{#2}}
\newcommand{\harrapp}[3]{#1(#2;#3)}

@@

\newcommand{\id}{\textrm{id}}
\newcommand{\for}{\textrm{for}\,}
\newcommand{\when}{\textrm{when}\,}
\newcommand{\lett}{\textrm{let}\,}
%% \newcommand{\sort}{\,\,\textrm{sort}}
\newcommand{\Sort}{\textrm{Sort}}
\newcommand{\isadtctx}{\,\,\textrm{type context}}
\newcommand{\isavectx}{\,\,\textrm{cat context}}
%% \newcommand{\obctx}{\,\,\textrm{object context}}
\newcommand{\boundary}{\,\,\textrm{boundary}}
\newcommand{\prof}{\,\,\textrm{span}}
\newcommand{\subst}{\,\,\textrm{subst}}
\newcommand{\sigctx}{\,\,\textrm{sig-ctx}}
\newcommand{\sig}{\,\,\textrm{sig}}
\newcommand{\pipe}{\mathrel{|}}

@@

@@

\begin{figure}
  \begin{mathpar}
    \begin{array}{rcl}
      \textrm{Type} & A,B,C ::= &
      \sum_{X:A} B \pipe \prod_{X:A} B \pipe \Id A M N \\
      \Cat \pipe \varr{\cat C}{\cat D}\pipe \harr {\cat C} {\cat D}\pipe \pendallXdotY {\alpha:{\cat C}} R\\
      \textrm{Term} & L,M,N ::= & X \pipe \lambda X:M. N \pipe M\,N \pipe (M,N) \pipe \pi_1 M \pipe \pi_2 M\\
      &&\lceil \cat C \rceil \pipe \vlambda {\alpha:\cat C} a \pipe \hlambda {\alpha:\cat C}{\beta: \cat D} R \ldots \\
      \textrm{Categories} &\cat C, \cat D, \cat E ::= & M \pipe \cat C \times \cat D \pipe \cat 1\\
      \textrm{Objects} & a, b, c ::= & \alpha \pipe M a \ldots\\
      \textrm{Sets} & P,Q,R ::= & \ldots \\
      \textrm{Transformations} & s,t,u & ::= & \ldots \\
      \\
      \textrm{Type Context} & \Gamma ::= & \cdot \pipe \Gamma, X : A\\
      \textrm{Cat Context} & \Phi,\Psi ::= & \cdot \pipe \alpha : \cat C \pipe \Phi , x : P , \Psi\\
    \end{array}
  \end{mathpar}
  \caption{Simple Category Theory Syntactic Forms}
\end{figure}

\subsection{Judgmental Structure}

The following figures present the basic formation, primitive and
admissible equality rules of our type theory, an internal language for
virtual double categories with restrictions.
%
The type theory has 4 central concepts, which we will refer to
sometimes using type theoretic/set theoretic and at other times
category-theoretic terminology.
%
In type theoretic terminology, judgment theory has \emph{sorts} $\cat
C, \cat D, \cat E$, and \emph{types} $A,B,C$ which have a given sort
and are also \emph{parameterized} by an object of some other sort, and
so can also be thought of as \emph{functions} with input and output
sorts.
%
Next there are \emph{judgments} $P,Q,R,S$ which are parameterized by
two variables.
%
In $\alpha^o:\cat C;\beta : \cat D \pipe R$, we say $R$ depends on
$\alpha$ \emph{contravariantly} and in $\beta$ \emph{covariantly}.
%
Finally, we have \emph{terms} $s,t,u$ that are elements of some
judgment $R$ but are also parameterized by elements of some other
judgments which are given by a \emph{context} $\Phi$, which is a
``compatible string'' of judgments where the covariant variable of one
judgment is the same as the contravariant variable of the next.
%
Since these terms are parameterized by other terms, we can see these
as \emph{inference rules} for the judgments.

At other times we will want to think of these using set-theoretic
terminology, in which case we have sets $\cat C,\cat D, \cat E$,
functions/elements $A,B,C$, \emph{spans} $R,\ldots$ and morphisms of
spans $s,t,u$.
%
Finally when we develop multicategory theory we will use
category-theoretic terminology, in which the structure is most
familiar and we have categories, objects/functors,
\emph{profunctors}/bimodules and 2-cells/homomorphisms of bimodules.

Figure\ref{fig:formation} introduces the formation rules for the basic
constructs. Sorts give a syntax for objects of the double
category. Types have an output sort and a sorted input variable, and
model vertical arrows. Next, judgments have two sorted variables, one
contravariant and one covariant, which can be the same
variable. Judgments model the horizontal arrows. Next are contexts,
whose formation rule is the same as judgments, and which give a syntax
for ``compatible strings'' of judgments. Then terms have an input
context and an output judgment, which must have the same covariant and
contravaiant variables, modeling ``globular'' 2-cells. Finally, we
have substitutions for terms, which are typed with respect to an input
context, an output context and two types, forming a kind of
square. These model the ``compatible string'' of 2-cells that arise in
the composition laws.

\begin{figure}
  \begin{mathpar}
    \inferrule
        {}
        {\Gamma \isadtctx}

    \inferrule
    {\Gamma \isadtctx}
    {\Gamma \vdash \Xi \boundary}

    \inferrule
    {\Gamma \isadtctx \and \Gamma \vdash \Xi \boundary}
    {\Gamma \pipe \Xi \vdash S \isaSet}

    \inferrule*[right=Object/Type-formation]
    {\Gamma \vdash \cat C : \Sort \and \Gamma \vdash \cat D : \Sort}
    {\Gamma \pipe \alpha : \cat C \vdash A : \cat D}\\

    \inferrule*[right=Context-formation]
    {\Gamma \isadtctx}
    {\Gamma \vdash \Phi \isavectx}

    \inferrule*[right=Term formation]
    {\Gamma \vdash \Phi \isavectx\and
     \Gamma\pipe \underline \Phi \vdash R \isaSet}
    {\Gamma \pipe \Phi \vdash M : R}

@@

Next, Figure~\ref{fig:basic-rules} gives the basic structural rules of
judgment theory, which model composition and restrictions in a virtual
double category. First, we have varialbes and substitution for types,
modeling identity and composition of vertical arrows. Next, we have
restriction of a judgment to more specific types, which is given by
substitution of types for the judgment's parameter variables. Next, we
have rules for constructing an ``empty'' context, which is just a

@@

    \inferrule
    {\Gamma \isadtctx \and \Gamma\pipe \cdot \vdash A \isaSet}
    {\Gamma , x : A \isadtctx}\\

    \inferrule{}{\Gamma \vdash \cdot \boundary}

    \inferrule{\Gamma \vdash C : \Sort}{\Gamma \vdash \alpha: C \boundary}

    \inferrule{\Gamma \vdash C : \Sort\and\Gamma\vdash D : \Sort}{\Gamma \vdash \alpha: C;\beta:D \boundary}\\

    \inferrule{}{\Gamma \vdash \cdot \isavectx}

    \inferrule{\Gamma \vdash C : \Sort}{\Gamma \vdash \alpha: C \isavectx}

    \inferrule
    {\Gamma \vdash \Phi \isavectx\and
    \Gamma \vdash \Psi \isavectx \and
    \Gamma \pipe d^+\Phi, d^- \Psi \vdash R \isaSet}
    {\Gamma \vdash \Phi, x:R, \Psi \isavectx}

    (\Phi,x:R,\Psi),y:S,\Sigma = \Phi,(x:R,\Psi,y:S,\Sigma)
  \end{mathpar}
  \begin{align*}
    \underline \cdot &= \cdot\\
    \underline {\alpha:C} &= \alpha:C\\
    \underline{\Phi,x:R,\Psi} &= d^-\Phi; d^+{\Psi}
  \end{align*}
  \begin{align*}
    d^{\pm}(\cdot) &\quad\text{undefined}\\
    d^{\pm}(\alpha:C) &= \alpha : C\\
    d^{-}(\Phi,x:R,\Psi) &= d^-\Phi\\
    d^{+}(\Phi,x:R,\Psi) &= d^-\Psi
  \end{align*}
  \caption{Basic Rules for Contexts}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \inferrule
    {}
    {\Gamma\pipe \cdot \vdash \Sort \isaSet}

    \inferrule
    {\Psi = \Psi_1 \jnctx \Psi_M \jnctx \Psi_2\and
      \Gamma \vdash \phi_1 : \Psi_1 \to \Phi_1\and
      \Gamma \pipe \Psi_M \vdash M : R[d^+\phi_1;d^-\phi_2] \and
      \Gamma \vdash \phi_2 : \Psi_2 \to \Phi_2
      }
    {\Gamma \vdash \phi_1,M/x,\phi_2 : \Psi \to \Phi_1,x:R,\Phi_2}
  \end{mathpar}
  \caption{Basic Rules of Judgment Theory}
  \label{fig:basic-rules}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \inferrule
    {\Gamma \vdash \Phi \isavectx \and
    \Gamma \vdash \Psi \isavectx \and
    d^+\Phi = d^- \Psi}
    {\Gamma \vdash \Phi \jnctx \Psi \isavectx}

Now that we have established a judgmental structure, we move on to the
multiplicative connectives, starting with the easier negative
connectives and then the more difficult positives.

\subsection{Left and Right Hom of Profunctors}

@@

>>>>

\usepackage{tikz-cd}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{stmaryrd}
\usepackage{mathpartir}
\usepackage{bbm}
\usepackage{cleveref}

@@

\begin{document}
%% \newtheorem{theorem}{Theorem}
%% \newtheorem{lemma}{Lemma}
%% \newtheorem{construction}{Construction}
%% \newtheorem{definition}{Definition}
\newcommand{\pto}{\nrightarrow}
\newcommand{\pfrom}{\nleftarrow}
\newcommand{\vcat}{\mathcal}
\newcommand{\cat}{\mathbbm}
\newcommand{\isaSet}{\,\,\textrm{Set}}
\newcommand{\isaTy}{\,\,\textrm{Type}}
\newcommand{\isaCat}{\,\,\textrm{Cat}}
\newcommand{\Set}{\textrm{Set}}
\newcommand{\vtkmnd}{\mathbb{K}\text{Mod} (\vcat{V},T)}
\newcommand{\rmod}{\text{RMod}}
\newcommand{\lmod}{\text{LMod}}
\newcommand{\varr}[2]{\text{Fun}{#1}{#2}}
\newcommand{\harr}[2]{\text{Rel}{#1}{#2}}
\newcommand{\harrapp}[3]{#1(#2;#3)}

@@

\newcommand{\id}{\textrm{id}}
\newcommand{\for}{\textrm{for}\,}
\newcommand{\when}{\textrm{when}\,}
\newcommand{\lett}{\textrm{let}\,}
%% \newcommand{\sort}{\,\,\textrm{sort}}
\newcommand{\Sort}{\textrm{Sort}}
\newcommand{\isadtctx}{\,\,\textrm{type context}}
\newcommand{\isavectx}{\,\,\textrm{trans. context}}
%% \newcommand{\obctx}{\,\,\textrm{object context}}
\newcommand{\boundary}{\,\,\textrm{set context}}
\newcommand{\prof}{\,\,\textrm{span}}
\newcommand{\subst}{\,\,\textrm{subst}}
\newcommand{\sigctx}{\,\,\textrm{sig-ctx}}
\newcommand{\sig}{\,\,\textrm{sig}}
\newcommand{\pipe}{\mathrel{|}}

@@

\newcommand{\Cat}{\textrm{Cat}}
\newcommand{\Id}[3]{\textrm{Id} #1\,#2\,#3}
\newcommand{\vlambda}[1]{\lambda^{F}{#1}.}
\newcommand{\hlambda}[2]{\lambda^{P}{#1};{#2}.}

@@

The syntactic forms of Simple Category Theory (SCT) are given in
\cref{fig:syntax}.
%
First, we have the types and terms of the logic over our category
theory. Types include the usual dependent sum, product and identity
types of dependent type theory, as well as several forms that allow us
to abstract over the category theoretic constructions: a type $\Cat$
of categories, a type $\varr{\cat C}{\cat D}$ for functors from $\cat
C$ to $\cat D$, a type $\harr{\cat C}{\cat D}$ for profunctors
contravariant in $\cat C$ and covariant in $\cat D$, and finally a
type $\pendallXdotY {\alpha:{\cat C}} R$ that, when combined with
other constructions on profunctors, allows for abstraction over
arbitrary natural transformations.
%
Terms $L,M,N$ describe the elements of our logic's types, and include
the usual constructions as well as forms for defining categories,
functors, profunctors and natural transformations.
%
Next, we have a syntax for categories, which are written using a
blackboard font $\cat C, \cat D, \cat E$. Categories include terms $M$
of the category type (allowing for abstraction), in addition to
primitives for product and unit categories.
%
Next, we have a syntax for objects of a category, which, because they
are typed with respect to an input variable, provide a syntax for
functors.
%
The objects we can define are object variables, application of a
functor $M a$, and pairs of objects (for the product category) and the
unique object of the unit category.
%
Next, $P,Q,R$ stand for \emph{sets}, but similar to objects, they are
always typed with respect to a contravariant and covariant variable,
and so denote profunctors.
%
These include first the multiplicatives: the hom set
$\punitinXfromYtoZ {\cat C} a b$ representing the morphisms from $a$
to $b$ in $\cat C$, a tensor product $\tensorexistsXwithYandZ
{\beta:\cat D} P Q$, and two kinds of ``ordered function types''
$\homrallXYtoZ \beta P Q$ and $\homlallXYtoZ \alpha R S$. 
%
We also include application of a profunctor term $M$ to two objects,
and cartesian product and unit sets.
%
Next, we have transformations $s,t,u$, which are typed with respect
to (0 or more) input and (one) output set.

Finally, we have three forms of contexts, type contexts $\Gamma$, set
contexts $\Xi$ and transformation contexts $\Phi$. A type context
$\Gamma$ is the familiar notion of context from dependent type
theory. We use uppercase $X,Y,Z$ for term variables to distinguish
them from set variables. Every set expression $P$ must depend on an
object of one category covariantly and an object of one category
contravariantly, however these are allowed to be the same
object. These two scenarios are captured by the notion of set context
$\Xi$. The set context $\alpha:\cat C; \beta: \cat D$ gives the common
case where the contravariant variable $\alpha$ is different from the
covariant variable $\beta$, and the set context $\alpha: \cat C$ alone
gives the case where the profunctor depends on the same variable both
covariantly and contravariantly. Note that we write object variables
as greek letters $\alpha, \beta$. Transformation contexts $\Phi$
describe the inputs to a natural transformation. The input variables
of a natural transformation $x: R$ stand for elements of sets, but as
just described, sets always have covariant and contravariant
dependencies. For this reason, transformation contexts include
category variables $\alpha: \cat C$ interleaved with the set variables
$x : R$, a typical context $\Phi$ would look like $\alpha_1:\cat C_1,
x_1:R_1, \alpha_2: \cat C_2, x_2 : R_2, \alpha_3: \cat C_3$, where
$R_1$ is parameterized by $\alpha_1$ contravariantly and $\alpha_2$
covariantly and $R_2$ is similarly parameterized by
$\alpha_2;\alpha_3$. Note that we consider transformation contexts up
to associativity in that $(\Phi, x:R, \Psi), y : S, \Sigma$ should be
considered equal to $\Phi, x:R, (\Psi, y: S, \Sigma)$.  In this
calculus there is no single ``empty'' transformation context, but
rather the empty context is one that consists of a single category
variable $\alpha:\cat C$ and no set variables. Each form of context
has a corresponding form of substitution, denoted using the lower-case
form of the same greek letters.

\begin{figure}
  \begin{mathpar}
    \begin{array}{rccl}
      \textrm{Type} & A,B,C & ::= &
      \sum_{X:A} B \pipe \prod_{X:A} B \pipe \Id A M N \\
      &&& \Cat \pipe \varr{\cat C}{\cat D}\pipe \harr {\cat C} {\cat D}\pipe \pendallXdotY {\alpha:{\cat C}} R\\
      \textrm{Term} & L,M,N & ::= & X \pipe \lambda X:M. N \pipe M\,N \pipe (M,N) \pipe \pi_1 M \pipe \pi_2 M\\
      &&&\lceil \cat C \rceil \pipe \vlambda {\alpha:\cat C} a \pipe \hlambda {\alpha:\cat C}{\beta: \cat D} R \ldots \\
      \textrm{Categories} &\cat C, \cat D, \cat E & ::= & M \pipe \cat C \times \cat D \pipe \cat 1\\
      \textrm{Objects} & a, b, c & ::= & \alpha \pipe M a \pipe (a,b) \pipe () \ldots\\
      \textrm{Sets} & P,Q,R & ::= & \punitinXfromYtoZ {\cat C} a b \pipe \tensorexistsXwithYandZ {\beta} P Q \pipe \homrallXYtoZ \beta P Q \pipe \homlallXYtoZ \alpha R S \\
      &&& M(a,b) \pipe 1 \pipe P \times Q\\
      \textrm{Transformations} & s,t,u & ::= & \ldots \\
      \\
      \textrm{Type Context} & \Gamma, \Delta & ::= & \cdot \pipe \Gamma, X : A\\
      \textrm{Category Context} & \Xi & ::= & \alpha:\cat C \pipe \alpha:\cat C;\beta: \cat D\\
      \textrm{Set Context} & \Phi,\Psi & ::= & \alpha : \cat C \pipe \Phi , x : P , \Psi\\
      \textrm{Term Substitution} & \gamma, \delta & ::= & \cdot \pipe \gamma, M/X\\
      \textrm{Object Substitution} & \xi, \zeta & ::= & a/\alpha \pipe a/\alpha;b/\beta\\
      \textrm{Transformation Substitution} & \phi, \psi & ::= & a/\alpha \pipe \phi,s/x, \psi      
    \end{array}
  \end{mathpar}
  \caption{Simple Category Theory Syntactic Forms}
  \label{fig:syntax}
\end{figure}

Next, \cref{fig:formation} shows the \emph{formation} rules for these
various syntactic forms. All other forms are typed with respect to a
well-formed type context $\Gamma$. Objects are typed additionally with
an input object variable $\alpha : \cat C$ and an output category
$\cat D$; in the semantics these objects are modeled as functors. A
set $S$ is typed with respect to a set context $\Xi$ to describe its
covariant/contravariant dependence on some input objects. These are
semantically modeled as profunctors. A transformation $s$ has a
context of transformation variable and a single set output. To be well
formed, the context and set must be parameterized by the same
contravariant and covariant object variables. To ensure this, we use a
coercion operation $\underline \Phi$ that erases everything in the
context but the left-most and right-most object variables.

\begin{figure}
  \begin{mathpar}
    \inferrule*[right=TyCtxForm]
        {}
        {\Gamma \isadtctx}

    \inferrule*[right=Cat-form]
    {\Gamma \isadtctx}
    {\Gamma\vdash \cat C \isaCat}

    \inferrule*[right=Ob-form]
    {\Gamma \vdash \cat C\isaCat\and \Gamma \vdash \cat D \isaCat}
    {\Gamma \pipe \alpha : \cat C \vdash a : \cat D}\\

    \inferrule*[right=SetCtxForm]
    {\Gamma \isadtctx}
    {\Gamma \vdash \Xi \boundary}

    \inferrule*[right=SetForm]
    {\Gamma \isadtctx \and \Gamma \vdash \Xi \boundary}
    {\Gamma \pipe \Xi \vdash S \isaSet}

    \inferrule*[right=TransCtxForm]
    {\Gamma \isadtctx}
     {\Gamma \vdash \Phi \isavectx}

    \inferrule*[right=TransFormation]
    {\Gamma \vdash \Phi \isavectx\and
     \Gamma\pipe \underline \Phi \vdash R \isaSet}
    {\Gamma \pipe \Phi \vdash s : R}

@@

%% The following figures present the basic formation, primitive and
%% admissible equality rules of our type theory, an internal language for
%% virtual double categories with restrictions.
%% %
%% The type theory has 4 central concepts, which we will refer to
%% sometimes using type theoretic/set theoretic and at other times
%% category-theoretic terminology.
%% %
%% In type theoretic terminology, judgment theory has \emph{sorts} $\cat
%% C, \cat D, \cat E$, and \emph{types} $A,B,C$ which have a given sort
%% and are also \emph{parameterized} by an object of some other sort, and
%% so can also be thought of as \emph{functions} with input and output
%% sorts.
%% %
%% Next there are \emph{judgments} $P,Q,R,S$ which are parameterized by
%% two variables.
%% %
%% In $\alpha^o:\cat C;\beta : \cat D \pipe R$, we say $R$ depends on
%% $\alpha$ \emph{contravariantly} and in $\beta$ \emph{covariantly}.
%% %
%% Finally, we have \emph{terms} $s,t,u$ that are elements of some
%% judgment $R$ but are also parameterized by elements of some other
%% judgments which are given by a \emph{context} $\Phi$, which is a
%% ``compatible string'' of judgments where the covariant variable of one
%% judgment is the same as the contravariant variable of the next.
%% %
%% Since these terms are parameterized by other terms, we can see these
%% as \emph{inference rules} for the judgments.

%% At other times we will want to think of these using set-theoretic
%% terminology, in which case we have sets $\cat C,\cat D, \cat E$,
%% functions/elements $A,B,C$, \emph{spans} $R,\ldots$ and morphisms of
%% spans $s,t,u$.
%% %
%% Finally when we develop multicategory theory we will use
%% category-theoretic terminology, in which the structure is most
%% familiar and we have categories, objects/functors,
%% \emph{profunctors}/bimodules and 2-cells/homomorphisms of bimodules.

%% Figure\ref{fig:formation} introduces the formation rules for the basic
%% constructs. Sorts give a syntax for objects of the double
%% category. Types have an output sort and a sorted input variable, and
%% model vertical arrows. Next, judgments have two sorted variables, one
%% contravariant and one covariant, which can be the same
%% variable. Judgments model the horizontal arrows. Next are contexts,
%% whose formation rule is the same as judgments, and which give a syntax
%% for ``compatible strings'' of judgments. Then terms have an input
%% context and an output judgment, which must have the same covariant and
%% contravaiant variables, modeling ``globular'' 2-cells. Finally, we
%% have substitutions for terms, which are typed with respect to an input
%% context, an output context and two types, forming a kind of
%% square. These model the ``compatible string'' of 2-cells that arise in
%% the composition laws.

%% Next, Figure~\ref{fig:basic-rules} gives the basic structural rules of
%% judgment theory, which model composition and restrictions in a virtual
%% double category. First, we have varialbes and substitution for types,
%% modeling identity and composition of vertical arrows. Next, we have
%% restriction of a judgment to more specific types, which is given by
%% substitution of types for the judgment's parameter variables. Next, we
%% have rules for constructing an ``empty'' context, which is just a
In typical natural-deduction style, all syntactic forms typed in
context admit an action of substitution. For terms, this is the
ordinary type theoretic substitution. Naturally objects $\alpha:\cat C
\vdash a : \cat D$ can be substituted for object variables $\beta :
\cat D$ in other objects. We can also substitute objects into
\emph{sets}. For instance, if we have a hom set in a category
$\punitinXfromYtoZ {\cat C} \alpha {\alpha'}$, and we can define
objects $a$ and $a'$ in the category $\cat C$, then we can substitute
those in to form the hom set between $a$ and $a'$: $\punitinXfromYtoZ
{\cat C} a {a'}$. This semantically is a ``restriction'' of a
profunctor along two functors, and modeling this operation simply as a
substitution considerably simplifies reasoning using profunctors.
%
Finally we have the action of substitution on natural
transformations. A natural transformation is indexed by set variables
$x:R$, whose sets are each parameterized by a contravariant and a
covariant category variable $\alpha;\beta$. We can think of natural
transformations as \emph{polymorphic} in the categories involved, and
so when we make a transformation substitution, we substitute not just
the set variables, but we also \emph{instantiate} the polymorphic
category variables with objects.

\cref{fig:contexts} details the typing rules for contexts and the
typing for variables and the admissible actions of substitution (the
definition of substitution is straightforward and found in the
appendix). the rules for term and object variables are as usual.
%
An object substitution into a single object context $\alpha:\cat C$
consists of an object $a : \cat C$ parameterized by an object of
another category.
%
An object substitution into a pair of co- and contra-variant variables
$\alpha:\cat C;\beta:\cat D$ is a pair of an object substituted for
$\alpha$ and one for $\beta$, each typed with respect to the
appropriate variable of an object context $\Xi$.
%
Note that this allows for $\Xi$ to consist of either two variables or
a single variable.
%
When
constructing a transformation context $\Phi, x : R, \Psi$, we need to
ensure that the contravariant dependence of $R$ matches the covariant
dependence of $\Phi$ and vice-versa for $\Psi$. We define this with
auxiliary metafunctions $d^+,d^-$ which extract the covariant and
contravariant object variable of a transformation context.
%
The rules for transformation variables and substitutions define the
use of variables to be \emph{linear} (each variables is used exactly
once) and \emph{ordered}, they occur in the term in the same order
that they appear in the context.
%
The rule for a transformation variables says a variable $x$ can only
be used when it is the sole variable in the context.
%
A transformation subsitution into an ``empty'' transformation context
$\alpha: \cat C$ is the same as an object substitution into
$\alpha:\cat C$.
%
The rule for constructing a transformation substitution into a
non-empty context $\Phi_1,x:R,\Phi_2$ says we can construct a
substitution by constructing substitutions $\phi_1,\phi_2$ for
$\Phi_1$ and $\Phi_2$ as well as a transformation $s$ into an
instantiation of $R$ where the input contexts come from splitting the
input context $\Psi$ into three components, and ensuring that the
instantiation of $R$ agrees with how $\phi_1,\phi_2$ instantiate the
variables. This uses two new metafunctions. First, $\Phi \jnctx \Psi$
is an operation for concatenating two contexts when $d^+(\Phi) =
d^-(\Psi)$. Next, we extend the boundary maps $d^\pm$ to substitutions
to extract their object substitutions.

@@

    \inferrule
    {\Gamma \isadtctx \and \Gamma \vdash A \isaTy}
    {\Gamma , X : A \isadtctx}\\

    \inferrule
    {X : A \in \Gamma}
    {\Gamma \vdash X : A}

    \inferrule
    {}
    {\alpha: \cat C \vdash \alpha: \cat C}

    \inferrule{\Gamma \vdash \cat C : \isaCat}{\Gamma \vdash \alpha: \cat C \boundary}

    \inferrule{\Gamma \vdash \cat C : \isaCat\and
      \Gamma\vdash \cat D : \isaCat}{\Gamma \vdash \alpha: \cat C;\beta:\cat D \boundary}\\

    \inferrule*
        {\Gamma \pipe \alpha':\cat C' \vdash A : \cat C}
        {\Gamma \vdash a/\alpha : \alpha':\cat C' \to \alpha:\cat C}

    \inferrule*
    {\Gamma \pipe d^-\Xi \vdash a : \cat C\and
     \Gamma \pipe d^+\Xi \vdash b : \cat D
    }
    {\Gamma \vdash (a/\alpha;b/\beta) : \Xi \to (\alpha:\cat C;\beta:\cat D)}\\

    \inferrule
        {\Gamma \vdash \cat C : \isaCat}
        {\Gamma \vdash \alpha: \cat C \isavectx}

    \inferrule
    {\Gamma \vdash \Phi \isavectx\and
    \Gamma \vdash \Psi \isavectx \and
    \Gamma \pipe d^+\Phi; d^- \Psi \vdash R \isaSet}
    {\Gamma \vdash \Phi, x:R, \Psi \isavectx}

    \inferrule
    {}
    {\Gamma \pipe \alpha:\cat C, x : R, \beta: \cat D \vdash x : R}

    \inferrule
    {\Psi = \Psi_1 \jnctx \Psi_M \jnctx \Psi_2\and
      \Gamma \vdash \phi_1 : \Psi_1 \to \Phi_1\and
      \Gamma \pipe \Psi_s \vdash s : R[d^+\phi_1;d^-\phi_2] \and
      \Gamma \vdash \phi_2 : \Psi_2 \to \Phi_2
      }
    {\Gamma \vdash \phi_1,t/x,\phi_2 : \Psi \to \Phi_1,x:R,\Phi_2}
  \end{mathpar}
  \begin{align*}
    \underline{\alpha:C} &= \alpha:C\\
    \underline{\Phi,x:R,\Psi} &= d^-\Phi; d^+{\Psi}
  \end{align*}
  \begin{align*}
    d^{\pm}(\alpha:\cat C) &= \alpha : \cat C\\
    d^{-}(\Phi,x:R,\Psi) &= d^-\Phi\\
    d^{+}(\Phi,x:R,\Psi) &= d^-\Psi
  \end{align*}
    \begin{align*}
    d^{\pm}(a/\alpha) &= a/\alpha\\
    d^{-}(\phi,t/x,\psi) &= d^-\phi\\
    d^{+}(\phi,t/x,\psi) &= d^+\psi\\
  \end{align*}
  \caption{Basic Rules of Judgment Theory}
  \label{fig:contexts}
\end{figure}

Now that we have established a judgmental structure, in typical type
theoretic fashion, our type constructs follow naturally as
internalizing the judgments.

\subsection{Negative Multiplicatives}

@@

TODO: define substitution
\begin{figure}
  \begin{mathpar}
    \inferrule
    {}
    {\Gamma\pipe \cdot \vdash \Sort \isaSet}

    \inferrule
    {\Psi = \Psi_1 \jnctx \Psi_M \jnctx \Psi_2\and
      \Gamma \vdash \phi_1 : \Psi_1 \to \Phi_1\and
      \Gamma \pipe \Psi_M \vdash M : R[d^+\phi_1;d^-\phi_2] \and
      \Gamma \vdash \phi_2 : \Psi_2 \to \Phi_2
      }
    {\Gamma \vdash \phi_1,M/x,\phi_2 : \Psi \to \Phi_1,x:R,\Phi_2}
  \end{mathpar}
  \caption{Substitutions in SCT}
  \label{fig:basic-rules}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \inferrule
    {}
    {\cdot : \cdot \to \cdot}

    \inferrule
    {\gamma : \Delta \to \Gamma \and \Delta \vdash M : A[\gamma]}
    {\gamma,M/X : \Delta \to \Gamma,X:A}

    \inferrule
    {\Gamma \vdash \Phi \isavectx \and
    \Gamma \vdash \Psi \isavectx \and
    d^+\Phi = d^- \Psi}
    {\Gamma \vdash \Phi \jnctx \Psi \isavectx}
